### 2014-04-29 text/Chapter03.tex
Structural data-flow (Chapter 3)

\subsection{Data-flow languages}
The Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an 
operation that accesses a context. In case of arrays, the context is neighbourhood of a current
location in the array specified by a cursor. In this section, we make the example more concrete,
using a simpler and better studied programming model, data-flow languages.

@@

When discussing data-flow languages in the previous section, we said that the context provides 
past values of variables. This can be viewed as a flat contextual property (the context needs
to keep all past values), but we can also view it as a structural property. Consider the following
example:
%
\begin{equation*}
\kvd{let}~\ident{offsetZip} = 0~\kvd{fby}~(\ident{left} + \kvd{prev}~\ident{right})
\end{equation*}
%
The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
To evaluate a current value of the stream, we need the current value of \ident{left} and one past
value of \ident{right}.

As mentioned earlier, a static analysis for data-flow computations could calculate how many past 
values must be cached. This can be done as a \emph{flat} coeffect analysis that produces just a 
single number for each function. However, we can design a more precise \emph{structural} analysis
and track the number of required elements for individual variables.

\paragraph{Dataflow coeffects.} The coeffect system for dataflow is
similar to the one for bounded reuse in that it tracks a vector
of numbers $\aclrd{R}$ as part of the context
$\coctx{\Gamma}{\aclrd{R}}$. Here, coeffects represent the maximal number of
past values (\emph{causality depth}) required for a variable.

Weakening, exchange, abstraction and sub-coeffecting are the same as in bounded linear
coeffects, but the remaining rules differ. In Figure~\ref{fig:dataflow-coeff},
accessed variables (\emph{var}) are annotated with
$0$ meaning that no past value is required (only the
current one). The (\emph{prev}) rule crates caching requirements --
it increments the number of required values for all variables
used in $e$ using scalar-vector addition.

Application and contraction have the same structure as before, 
but use different operators. If two variables 
are contracting, requiring $\cclrd{s}$ and $\cclrd{t}$ past
values, then overall we need at most
$\ident{max}(\cclrd{s}, \cclrd{t})$ past values (\emph{contr}). That is, 
two caches are combined with the maximum of the two requirements, 
which satisfy the smaller requirements.

In (\emph{app}), the function requires $\cclrd{t}$ past values of
its parameter.  This means $\cclrd{t}$ past values of
$e_2$ are needed which in turn requires $\aclrd{S}$ past values of its free
variables $\Gamma_2$. Thus, we need $\cclrd{t}+\aclrd{S}$ past values of $\Gamma_2$ to perform the
call (\eg{}, we need $1+\aclrd{S}$ values to
get $1$ past value of the input $\sigma$, $2+\aclrd{S}$
values to get $2$ past values of $\sigma$, \emph{etc.}).

\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
and one bound:
\[
\dfrac
  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
\]
The expression always requires the previous value of $y$ and adds it to
a previous value of the parameter $x$. Evaluating the value of the
argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past values of $y$
and so the overall requirement is $3$ past values:
\[
\inference
  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) \quad  \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~x) : \mathbb{Z} }
{\inference
  { \coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~x)) : \mathbb{Z} }
  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
\]
The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then 
(\emph{contr}) to take the maximum, showing three past values are sufficient. Reducing 
the expression by substitution we get $\kvd{prev}~(y+(\kvd{prev}~(\kvd{prev}~y)))$. 
Semantically, this performs stream lookups $y[1] + y[3]$ where the indices are the 
number of enclosing $\kvd{prev}$s.% expressions.

We previously used dataflow as an example of
coeffects~\cite{petricek2013coeffects}, but tracked caching
requirements on the whole context. The system outlined here is more
powerful and practically useful, with finer-grained coeffects tracking
caching requirements per-variable.

% --------------------------------------------------------------------------------------------------
% variable annotated with 0 (not 1)
% sub and abs omitted
% app uses +, not \ast
% let is default

@@

>>>>

\subsection{Data-flow languages}
\label{sec:applications-flat-dataflow}

The Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an 
operation that accesses a context. In case of arrays, the context is neighbourhood of a current
location in the array specified by a cursor. In this section, we make the example more concrete,
using a simpler and better studied programming model, data-flow languages.

@@

When discussing data-flow languages in the previous section, we said that the context provides 
past values of variables. In Section~\ref{sec:applications-flat-dataflow}, we tracked this as 
a \emph{flat} property, which gives us a system that keeps the same number of past values for
all variables. However, data-flow can also be adapted to a structural system which keeps the number 
of required past values individually for each variable. Consider the 
following example:
%
\begin{equation*}
\kvd{let}~\ident{offsetZip} = \ident{left} + \kvd{prev}~\ident{right}
\end{equation*}
%
The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
To evaluate a current value of the stream, we need the current value of \ident{left} and one past
value of \ident{right}. Flat system is not able to capture this level-of-detail and simply
requires $1$ past values of both streams in the variable context.

Turning a flat data-flow system to a structural data-flow system is a change similar to the one
between flat ans structural liveness. In case of liveness analysis, we included the flat system
only as an illustration (it is not practically useful).

For data-flow, the flat system is less precise, but still practically useful (simplicity may outweigh 
precision). As discussed in Section~X, the structural system is necessary when we allow arbitrary 
recursion and cannot (easily) determine the number of required values statically.

@@

\paragraph{Type system.} 
The type system in Figure~\ref{fig:applications-struct-df} annotates the variable context with a
vector of integers. This is similar as in the bounded reuse system, but the integers \emph{mean} a
different thing. Consequently, they are also calculated differently. We omit rules that are the
same for all structural coeffect systems (exchange, lambda abstraction).

In data-flow, we annotate both used variables (\emph{var}) and unused variables (\emph{weak}) with
$0$, meaning that no past values are required. This is the same as in flat data-flow, but different
from bounded reuse and liveness (where difference between using and not using a variable matters).
Primitive requirements are introduced by the (\emph{prev}) rule, which increments the annotation of
all variables in the context.

In flat data-flow, we identified sequential composition and point-wise composition as two primitive
operations that were used in the (flat) application. In the structural system, these are used in 
(\emph{app}) and (\emph{contr}), respectively. Thus application combines coeffect annotations using
$+$ and contraction using \emph{max}. This contrasts with bounded reuse, which uses $\ast$ and $+$,
respectively.

\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
and one bound. The (\emph{abs}) rule splits the annotations between the declaration-site and call-site
of the function:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
\end{equation*}
%
The expression always requires the previous value of $y$ and adds it to a previous value of the 
parameter $x$. Evaluating the value of the argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past 
values of $y$ and so the overall requirement for the (free) variable $y$ is $3$ past values. In 
order to use the contraction rule, we rename $y$ to $y'$ in the argument:
%
\begin{equation*}
\inference
  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} &
    \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~y') : \mathbb{Z} }
{\inference
  { \coctx{y\!:\!\mathbb{Z}, y'\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y')) : \mathbb{Z} }
  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
\end{equation*}
%
The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then (\emph{contr}) to take 
the maximum, showing three past values are sufficient.

Note that we get the same requirements when we perform $\beta$ reduction of the expression. 
Substituting the argument for $x$ yields the expression $\kvd{prev}~(y+(\kvd{prev}~(\kvd{prev}~y)))$. 
Semantically, this performs stream lookups $y[1]$ and $y[3]$ where the indices are the 
number of enclosing $\kvd{prev}$ constructs.

\paragraph{Semantics.}
To define the semantics of our structural data-flow language, we can use the same approach as when
adapting flat liveness to structural liveness. Rather than wrapping the whole context in some wrapper
(list or option type), we now wrap individual components of the product representing the variables
in the context.

The result is similar as the structure used for bounded reuse. The only difference is that, given
a variable annotated with $\cclrd{r}$, we need $1+\cclrd{r}$ values. That is, we need the current
value, followed by $\cclrd{r}$ past values:
%
\begin{equation*}
\begin{array}{l}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
  ~:~ (\tau^{(\cclrd{r_1}+1)}_1 \times \ldots \times \tau^{(\cclrd{r_n}+1)}_n) \rightarrow \tau\\
\sem{\tau_1 \xrightarrow{\cclrd{s}} \tau_2} ~=~ \tau_1^{(\cclrd{s}+1)} \rightarrow \tau_2
\end{array}
\end{equation*}
%
Despite the similarity with the semantics for bounded reuse, the values here \emph{represent}
different things. Rather than providing multiple copies of a value (out of which each can be
used just once), the pair provides past values (that can be reused and freely accessed).
To illustrate the behaviour we consider the semantics of the \kvd{prev} construct and of the
structural contraction rule:
%
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{prev}}
\begin{array}{l}
\sem{\coctx{\Gamma}{\alift{(\cclrd{s}_1 + 1), \ldots, (\cclrd{s}_n + 1)}} \vdash \kvd{prev}~e : \tau} = \\[-0.25em]
  \qquad\lambda ((x_{1,0}, \ldots, x_{1,{\cclrd{s}_1 + 1}}), \ldots, (x_{n,0}, \ldots, x_{n,{\cclrd{s}_n + 1}})) \rightarrow \\[-0.25em]
\qquad\qquad\sem{\coctx{\Gamma}{\alift{\cclrd{s}_1, \ldots, \cclrd{s}_n}} \vdash e : \tau}\\[-0.25em]
\qquad\qquad\qquad ((x_{1,0}, \ldots, x_{1,\cclrd{s}_1}), \ldots, (x_{n,0}, \ldots, x_{n,\cclrd{s}_n}))
\end{array}
\end{equation*}
%
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{contr}}
\begin{array}{l}
\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\textnormal{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} =\\[-0.25em] 
\qquad  \lambda(\mathbf{v_1}, (x_0, x_1, \ldots, x_{ \textnormal{max}(\cclrd{s},\cclrd{t}) }), \mathbf{v_2}) \rightarrow \\[-0.25em]
\qquad\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}\\[-0.25em]
\qquad\qquad\qquad (\mathbf{v_1}, (x_0, \ldots, x_{\cclrd{s}}), (x_0, \ldots, x_{\cclrd{t}} ), \mathbf{v_2})
\end{array}
\end{equation*}
%
In (\emph{prev}), the semantic function is called with an argument that stores values of $n$
variables, such that a variable $x_i$ has values ranging from $x_{i,0}$ to $x_{i,\cclrd{s}_i + 1}$.
Thus, there is one current value, followed by $\cclrd{s_i} + 1$ past values. The expression $e$
nested under $\kvd{prev}$ requires only $\cclrd{s_i}$ past values and so the semantics simply
drops the last value.

In the (\emph{contr}) rule, the semantic function receives $\emph{max}(\cclrd{s}, \cclrd{t})$ values
of a specific variable $x$. It needs to produce values for two separate variables, $y$ and $z$ that
require $\cclrd{s}$ and $\cclrd{t}$ past values. Both of these numbers are certainly smaller than 
(or equal to) the number of values available. Thus we simply take the first values. Unlike in the
contraction for BLL, the values are duplicated and the same values are used for both variables.

\paragraph{Summary.}
Two of the structural examples shown so far (liveness and data-flow) extend an earlier flat 
version of a similar system. We discuss this relation in general later. However, a flat system 
can generally be turned into a structural one -- although this only gives a useful system when
the flat version captures properties related to variables.

The data-flow example demonstrates that the a flat system can also be turned into structural 
system. In general, this only works for systems where lambda abstraction duplicates context
requirements (as in Figure~\ref{fig:applications-flat-liveness}).
